---
title: "Token-Aware Context Building"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Token-Aware Context Building}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Why context building matters

LLMs have finite context windows. A typical model accepts between 4,000 and
128,000 tokens, and every token you spend on context is a token you cannot
spend on the model's response. When building RAG applications, you often have
more retrieved content than fits in the window -- system prompts, retrieved
chunks, conversation history, and user instructions all compete for space.

Naive approaches (concatenate everything, hope it fits) fail in predictable
ways: either the prompt is truncated silently, or the model ignores information
buried in the middle of a long context. securecontext's `context_builder()`
solves this with a priority-based token budget that gives you explicit control
over what gets included and clear reporting on what gets dropped.

## How the context builder works

The builder follows a simple algorithm:

```
                    +------------------+
                    | Set token budget |
                    | (max_tokens)     |
                    +--------+---------+
                             |
                    +--------v---------+
                    | Sort items by    |
                    | priority (desc)  |
                    +--------+---------+
                             |
                +------------v------------+
                | For each item (highest  |
                | priority first):        |
                |                         |
                |  Estimate token count   |
                |         |               |
                |    Fits in budget?       |
                |    /           \         |
                |  YES           NO       |
                |   |             |        |
                | Include      Exclude    |
                | (deduct      (record    |
                |  tokens)      label)    |
                +------------+------------+
                             |
                    +--------v---------+
                    | Return:          |
                    | - context string |
                    | - included list  |
                    | - excluded list  |
                    | - total_tokens   |
                    +------------------+
```

Items with the highest priority numbers are included first. When the remaining
budget cannot accommodate the next item, that item (and all lower-priority
items) are excluded. The builder reports both lists, so you always know exactly
what the LLM will see and what it will not.

## Basic usage

Create a builder with a token budget, add content with priorities, and build
the final context string.

```{r}
library(securecontext)

cb <- context_builder(max_tokens = 100)

# Add content with different priorities (higher = included first)
cb <- cb_add(cb, "You are a helpful assistant.", priority = 10, label = "system")
cb <- cb_add(cb,
  "R is great for statistics and data visualization.",
  priority = 5, label = "retrieved_chunk_1"
)
cb <- cb_add(cb,
  "Python is popular for machine learning and web development.",
  priority = 4, label = "retrieved_chunk_2"
)
cb <- cb_add(cb,
  "Julia offers high performance for numerical computing workloads.",
  priority = 3, label = "retrieved_chunk_3"
)

result <- cb_build(cb)

cat("Assembled context:\n")
cat(result$context, "\n\n")
cat("Included:", paste(result$included, collapse = ", "), "\n")
cat("Excluded:", paste(result$excluded, collapse = ", "), "\n")
cat("Total tokens:", result$total_tokens, "\n")
```

With a 100-token budget, the system prompt (priority 10) is always included.
Retrieved chunks are then packed in priority order until the budget is
exhausted. The `$excluded` field tells you exactly which chunks were dropped,
which is valuable for debugging and logging in production agents.

## Priority design patterns

Choosing priority values is a design decision that depends on your application.
Here are common patterns:

Fixed tiers assign priorities by content type:

| Priority | Content type        | Rationale                              |
|---------:|:--------------------|:---------------------------------------|
| 10       | System prompt       | Defines agent behavior, always needed  |
| 7        | User's question     | The query must be visible to the model |
| 5        | Retrieved chunks    | Supporting evidence, best-effort       |
| 3        | Conversation history| Helpful but expendable                 |
| 1        | Disclaimers/footers | Include only if space permits           |

Dynamic priorities use retrieval scores directly. When you retrieve chunks
with cosine similarity scores, you can pass those scores as priorities so the
most relevant chunks are included first:

```{r}
cb <- context_builder(max_tokens = 500)

# System prompt gets highest priority -- always included
cb <- cb_add(cb, "You are an R expert.", priority = 10, label = "system")

# Retrieved chunks get decreasing priority by relevance score
hits <- retrieve(ret, "statistical models", k = 5)
for (i in seq_len(nrow(hits))) {
  chunk_text <- hits$id[i]  # or look up the original text
  cb <- cb_add(cb, chunk_text, priority = hits$score[i], label = hits$id[i])
}

result <- cb_build(cb)
cat("Included:", paste(result$included, collapse = ", "), "\n")
cat("Excluded:", paste(result$excluded, collapse = ", "), "\n")
cat("Total tokens:", result$total_tokens, "\n")
```

## Resetting between turns

In a multi-turn conversation, you typically rebuild the context for each turn --
the retrieved chunks change, the conversation history grows, and the system
prompt may be updated. Use `cb_reset()` to clear all items and reuse the same
builder without re-specifying the token budget.

```{r}
cb2 <- cb_reset(cb)
cb2 <- cb_add(cb2, "New system prompt.", priority = 10, label = "system_v2")
result2 <- cb_build(cb2)
cat("After reset -- included:", paste(result2$included, collapse = ", "), "\n")
```

This avoids the overhead of creating a new builder object each turn and makes
intent clear: this is a fresh context assembly for a new turn.

## The `context_for_chat()` shortcut

For the common case of "retrieve chunks and build a context string,"
`context_for_chat()` combines both steps in a single call. It retrieves the
top-k chunks from a retriever and packs them into a token-limited string.

```{r}
result <- context_for_chat(ret, "statistics", max_tokens = 2000)
cat(result$context)
```

Under the hood, this creates a context builder, adds each retrieved chunk with
its similarity score as the priority, and returns the built result. It is a
convenience wrapper; when you need more control (e.g., adding a system prompt
or conversation history alongside retrieved chunks), use the builder directly.
