---
title: "RAG-Enabled Agents"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RAG-Enabled Agents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Why Integrate Retrieval with Orchestration?

An LLM agent that can retrieve information from a knowledge base before
answering is fundamentally more capable than one that relies solely on its
training data. But retrieval alone is not enough -- you also need to decide
*when* to retrieve, *how much* context to include, and *where* to store what
the agent learns. These are orchestration concerns.

securecontext handles the retrieval side: document ingestion, embedding, vector
search, and token-aware context assembly. orchestr handles the orchestration
side: defining agent workflows as directed graphs, managing state between
nodes, and routing execution. Together, they let you build RAG agents where
retrieval is a first-class step in the workflow, not an afterthought bolted
onto a prompt.

This vignette shows how to connect the two packages using the memory adapter
pattern and graph-based retrieve-then-generate workflows.

For securecontext basics, see `vignette("securecontext")` and
`vignette("retrieval-workflows")`. For orchestr basics, see
`vignette("quickstart", package = "orchestr")`.

## Building a Knowledge Base

Start by creating documents, chunking them, and loading them into a retriever.
This is the same pipeline from `vignette("retrieval-workflows")`, repeated
here for completeness. Everything runs locally -- no external API calls
required.

```{r knowledge-base}
library(securecontext)
library(orchestr)
library(ellmer)

# Create documents from your corpus
docs <- list(
  document("R provides extensive facilities for statistical computing
and graphics. Linear models, time-series analysis, classification, and
clustering are all available out of the box.",
    metadata = list(source = "r-intro", topic = "statistics")),
  document("The tidyverse is a collection of R packages for data science.
Core packages include ggplot2, dplyr, tidyr, readr, and purrr.",
    metadata = list(source = "r-intro", topic = "tidyverse")),
  document("Shiny is an R package for building interactive web applications.
It combines R's analytical power with modern web UI components.",
    metadata = list(source = "r-intro", topic = "shiny"))
)

# Build embedder from the corpus
corpus <- vapply(docs, function(d) d@text, character(1))
embedder <- embed_tfidf(corpus)

# Create vector store and retriever
vs <- vector_store$new(dims = embedder@dims)
ret <- retriever(vs, embedder)
add_documents(ret, docs, chunk_strategy = "sentence")
```

## The `as_orchestr_memory()` Adapter

orchestr agents expect a memory backend with `get()` and `set()` methods.
securecontext's `knowledge_store` provides persistent key-value storage backed
by JSONL, but its interface does not match what orchestr expects out of the
box. The `as_orchestr_memory()` function bridges this gap.

The adapter pattern is simple: wrap a securecontext object in a thin layer that
exposes the interface another package expects. The underlying storage,
persistence, and search capabilities of the knowledge store are fully
preserved -- the adapter just translates method calls.

```{r memory-adapter}
# Create a persistent knowledge store
ks <- knowledge_store$new(path = "agent-memory.jsonl")

# Wrap it for orchestr
mem <- as_orchestr_memory(ks)

# The adapter exposes get/set -- the same interface orchestr expects
mem$set("user.name", "Alice")
mem$set("session.topic", "data analysis")
mem$get("user.name")
#> [1] "Alice"
```

The underlying JSONL file persists across R sessions, so agent memory survives
restarts. You can also access the knowledge store directly (via `ks`) to use
features like `$search()` and `$list()` that are not part of the orchestr
memory interface.

## Retrieval-in-the-Loop Graph

The core RAG pattern is: retrieve relevant context before each LLM call. With
orchestr's `graph_builder()`, you wire this as a two-node graph: a retrieval
node followed by an agent node.

The following diagram shows the flow:

```
  +----------+      +---------+      +-----+
  | retrieve | ---> |  agent  | ---> | END |
  +----------+      +---------+      +-----+
       |                  |
  Searches vector    Uses retrieved
  store, builds      context to
  token-limited      answer query
  context string     via LLM
```

The retrieval node runs the securecontext pipeline (retrieve + context build).
The agent node takes that context and passes it to an LLM alongside the user's
question. By separating these concerns into distinct graph nodes, each step is
independently testable and replaceable.

```{r rag-graph}
# Node 1: retrieve relevant chunks and build context
retrieve_node <- function(state, config) {
  query <- state$messages[[length(state$messages)]]

  # Retrieve and assemble token-limited context
  result <- context_for_chat(ret, query, max_tokens = 2000, k = 5)

  list(context = result$context)
}

# Node 2: LLM agent that uses the retrieved context
agent_node <- function(state, config) {
  context <- state$context %||% ""
  query <- state$messages[[length(state$messages)]]

  prompt <- paste0(
    "Use the following context to answer the question.\n\n",
    "Context:\n", context, "\n\n",
    "Question: ", query
  )

  chat <- chat_anthropic(system_prompt = "You are a helpful R assistant.")
  response <- chat$chat(prompt)

  list(messages = list(response))
}

# Wire the graph: retrieve -> agent -> END
schema <- state_schema(
  messages = "append:list",
  context = "character"
)

graph <- graph_builder(state_schema = schema)
graph$add_node("retrieve", retrieve_node)
graph$add_node("agent", agent_node)
graph$add_edge("retrieve", "agent")
graph$add_edge("agent", END)
graph$set_entry_point("retrieve")

rag_graph <- graph$compile()

# Run it
result <- rag_graph$invoke(list(
  messages = list("What packages are in the tidyverse?")
))
```

Every user query first passes through the retrieval node, which searches the
vector store and assembles a context string. The agent node then answers using
that context. Because the context builder enforces a token budget, the agent
node always receives a prompt that fits within the model's context window.

## Token Budget Management

When your knowledge base is large, retrieved chunks may exceed the LLM's context
window. The `context_builder()` controls this with a token budget and priorities.
For a detailed treatment of priority strategies and overflow behavior, see
`vignette("context-building")`.

Here is a practical example using retrieval scores as priorities:

```{r token-budget}
cb <- context_builder(max_tokens = 500)

# System prompt gets highest priority -- always included
cb <- cb_add(cb, "You are an R expert.", priority = 10, label = "system")

# Retrieved chunks get decreasing priority by relevance score
hits <- retrieve(ret, "statistical models", k = 5)
for (i in seq_len(nrow(hits))) {
  chunk_text <- hits$id[i]  # or look up the original text
  cb <- cb_add(cb, chunk_text, priority = hits$score[i], label = hits$id[i])
}

result <- cb_build(cb)
cat("Included:", paste(result$included, collapse = ", "), "\n")
cat("Excluded:", paste(result$excluded, collapse = ", "), "\n")
cat("Total tokens:", result$total_tokens, "\n")
```

The builder packs items in priority order until the budget is exhausted. Dropped
items are reported in `$excluded`, so you can log what was cut.

Use `cb_reset()` between turns to reuse the same builder:

```{r reset}
cb <- cb_reset(cb)
# Now add fresh content for the next turn
```

## Persistent Knowledge Store

For agents that need memory across sessions, `knowledge_store` persists to a
JSONL file. Combined with the orchestr memory adapter, this gives agents
durable recall -- the agent can remember user preferences, past queries, and
learned facts across restarts.

```{r persistent}
# Knowledge store persists to disk
ks <- knowledge_store$new(path = "long-term-memory.jsonl")

# Store structured facts
ks$set("user.preference", list(language = "R", theme = "dark"))
ks$set("session.2025-01-15", list(
  topic = "regression models",
  outcome = "built linear model for mtcars"
))

# Search by key pattern
ks$search("^session")
#> $`session.2025-01-15`
#> $`session.2025-01-15`$topic
#> [1] "regression models"

# Use in an orchestr agent via the adapter
mem <- as_orchestr_memory(ks)
```

## Putting It All Together

Here is a complete RAG agent that combines retrieval, token management, and
persistent memory in an orchestr graph. This represents a production-ready
pattern: documents are ingested into a local vector store, queries trigger
retrieval and context assembly, the LLM answers using grounded context, and
the agent persists what it learns for future sessions.

```{r full-example}
library(securecontext)
library(orchestr)
library(ellmer)

# --- Knowledge base ---
docs <- list(
  document("dplyr provides a grammar of data manipulation with verbs
like filter, select, mutate, summarise, and arrange."),
  document("ggplot2 implements the grammar of graphics. Build plots
layer by layer with aes(), geom_point(), geom_line(), and facet_wrap()."),
  document("tidyr helps tidy data with pivot_longer, pivot_wider,
separate, and unite.")
)

corpus <- vapply(docs, function(d) d@text, character(1))
embedder <- embed_tfidf(corpus)
vs <- vector_store$new(dims = embedder@dims)
ret <- retriever(vs, embedder)
add_documents(ret, docs, chunk_strategy = "sentence")

# --- Persistent memory ---
ks <- knowledge_store$new(path = tempfile(fileext = ".jsonl"))
mem <- as_orchestr_memory(ks)

# --- Graph nodes ---
retrieve_node <- function(state, config) {
  query <- state$messages[[length(state$messages)]]
  result <- context_for_chat(ret, query, max_tokens = 1500, k = 3)
  list(context = result$context)
}

agent_node <- function(state, config) {
  context <- state$context %||% ""
  query <- state$messages[[length(state$messages)]]

  prompt <- paste0("Context:\n", context, "\n\nQuestion: ", query)
  chat <- chat_anthropic(
    system_prompt = "Answer questions about R packages using the provided context."
  )
  response <- chat$chat(prompt)

  # Persist what we learned
  mem$set(paste0("query.", Sys.time()), query)

  list(messages = list(response))
}

# --- Build and run ---
schema <- state_schema(messages = "append:list", context = "character")
g <- graph_builder(state_schema = schema)
g$add_node("retrieve", retrieve_node)
g$add_node("agent", agent_node)
g$add_edge("retrieve", "agent")
g$add_edge("agent", END)
g$set_entry_point("retrieve")

rag <- g$compile()
result <- rag$invoke(list(
  messages = list("How do I reshape data from wide to long format?")
))
```

The retrieval node finds chunks about tidyr's `pivot_longer`, the context
builder fits them within the token budget, and the agent answers using that
grounded context. The query is also persisted to the knowledge store for
future reference.
