---
title: "RAG-Enabled Agents"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RAG-Enabled Agents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette shows how to combine securecontext and orchestr to build
retrieval-augmented generation (RAG) agents. securecontext handles document
ingestion, embedding, retrieval, and token-budgeted context assembly; orchestr
handles agent orchestration and graph-based workflows.

For securecontext basics, see `vignette("securecontext")` and
`vignette("retrieval-workflows")`. For orchestr basics, see
`vignette("quickstart", package = "orchestr")`.

## Building a Knowledge Base

Start by creating documents, chunking them, and loading them into a retriever.
Everything runs locally -- no external API calls required.

```{r knowledge-base}
library(securecontext)
library(orchestr)
library(ellmer)

# Create documents from your corpus
docs <- list(
  document("R provides extensive facilities for statistical computing
and graphics. Linear models, time-series analysis, classification, and
clustering are all available out of the box.",
    metadata = list(source = "r-intro", topic = "statistics")),
  document("The tidyverse is a collection of R packages for data science.
Core packages include ggplot2, dplyr, tidyr, readr, and purrr.",
    metadata = list(source = "r-intro", topic = "tidyverse")),
  document("Shiny is an R package for building interactive web applications.
It combines R's analytical power with modern web UI components.",
    metadata = list(source = "r-intro", topic = "shiny"))
)

# Build embedder from the corpus
corpus <- vapply(docs, function(d) d@text, character(1))
embedder <- embed_tfidf(corpus)

# Create vector store and retriever
vs <- vector_store$new(dims = embedder@dims)
ret <- retriever(vs, embedder)
add_documents(ret, docs, chunk_strategy = "sentence")
```

## The `as_orchestr_memory()` Adapter

securecontext's `knowledge_store` provides persistent key-value storage backed
by JSONL. The `as_orchestr_memory()` function wraps it so orchestr agents can
use it as their memory backend.

```{r memory-adapter}
# Create a persistent knowledge store
ks <- knowledge_store$new(path = "agent-memory.jsonl")

# Wrap it for orchestr
mem <- as_orchestr_memory(ks)

# The adapter exposes get/set -- the same interface orchestr expects
mem$set("user.name", "Alice")
mem$set("session.topic", "data analysis")
mem$get("user.name")
#> [1] "Alice"
```

The underlying JSONL file persists across R sessions, so agent memory survives
restarts.

## Retrieval-in-the-Loop Graph

The core RAG pattern is: retrieve relevant context before each LLM call. With
orchestr's `graph_builder()`, you wire this as a two-node graph: a retrieval
node followed by an agent node.

```{r rag-graph}
# Node 1: retrieve relevant chunks and build context
retrieve_node <- function(state, config) {
  query <- state$messages[[length(state$messages)]]

  # Retrieve and assemble token-limited context
  result <- context_for_chat(ret, query, max_tokens = 2000, k = 5)

  list(context = result$context)
}

# Node 2: LLM agent that uses the retrieved context
agent_node <- function(state, config) {
  context <- state$context %||% ""
  query <- state$messages[[length(state$messages)]]

  prompt <- paste0(
    "Use the following context to answer the question.\n\n",
    "Context:\n", context, "\n\n",
    "Question: ", query
  )

  chat <- chat_anthropic(system_prompt = "You are a helpful R assistant.")
  response <- chat$chat(prompt)

  list(messages = list(response))
}

# Wire the graph: retrieve -> agent -> END
schema <- state_schema(
  messages = "append:list",
  context = "character"
)

graph <- graph_builder(state_schema = schema)
graph$add_node("retrieve", retrieve_node)
graph$add_node("agent", agent_node)
graph$add_edge("retrieve", "agent")
graph$add_edge("agent", END)
graph$set_entry_point("retrieve")

rag_graph <- graph$compile()

# Run it
result <- rag_graph$invoke(list(
  messages = list("What packages are in the tidyverse?")
))
```

Every user query first passes through the retrieval node, which searches the
vector store and assembles a context string. The agent node then answers using
that context.

## Token Budget Management

When your knowledge base is large, retrieved chunks may exceed the LLM's context
window. The `context_builder()` controls this with a token budget and priorities.

```{r token-budget}
cb <- context_builder(max_tokens = 500)

# System prompt gets highest priority -- always included
cb <- cb_add(cb, "You are an R expert.", priority = 10, label = "system")

# Retrieved chunks get decreasing priority by relevance score
hits <- retrieve(ret, "statistical models", k = 5)
for (i in seq_len(nrow(hits))) {
  chunk_text <- hits$id[i]  # or look up the original text
  cb <- cb_add(cb, chunk_text, priority = hits$score[i], label = hits$id[i])
}

result <- cb_build(cb)
cat("Included:", paste(result$included, collapse = ", "), "\n")
cat("Excluded:", paste(result$excluded, collapse = ", "), "\n")
cat("Total tokens:", result$total_tokens, "\n")
```

The builder packs items in priority order until the budget is exhausted. Dropped
items are reported in `$excluded`, so you can log what was cut.

Use `cb_reset()` between turns to reuse the same builder:

```{r reset}
cb <- cb_reset(cb)
# Now add fresh content for the next turn
```

## Persistent Knowledge Store

For agents that need memory across sessions, `knowledge_store` persists to a
JSONL file. Combined with the orchestr memory adapter, this gives agents
durable recall.

```{r persistent}
# Knowledge store persists to disk
ks <- knowledge_store$new(path = "long-term-memory.jsonl")

# Store structured facts
ks$set("user.preference", list(language = "R", theme = "dark"))
ks$set("session.2025-01-15", list(
  topic = "regression models",
  outcome = "built linear model for mtcars"
))

# Search by key pattern
ks$search("^session")
#> $`session.2025-01-15`
#> $`session.2025-01-15`$topic
#> [1] "regression models"

# Use in an orchestr agent via the adapter
mem <- as_orchestr_memory(ks)
```

## Putting It All Together

Here is a complete RAG agent that combines retrieval, token management, and
persistent memory in an orchestr graph.

```{r full-example}
library(securecontext)
library(orchestr)
library(ellmer)

# --- Knowledge base ---
docs <- list(
  document("dplyr provides a grammar of data manipulation with verbs
like filter, select, mutate, summarise, and arrange."),
  document("ggplot2 implements the grammar of graphics. Build plots
layer by layer with aes(), geom_point(), geom_line(), and facet_wrap()."),
  document("tidyr helps tidy data with pivot_longer, pivot_wider,
separate, and unite.")
)

corpus <- vapply(docs, function(d) d@text, character(1))
embedder <- embed_tfidf(corpus)
vs <- vector_store$new(dims = embedder@dims)
ret <- retriever(vs, embedder)
add_documents(ret, docs, chunk_strategy = "sentence")

# --- Persistent memory ---
ks <- knowledge_store$new(path = tempfile(fileext = ".jsonl"))
mem <- as_orchestr_memory(ks)

# --- Graph nodes ---
retrieve_node <- function(state, config) {
  query <- state$messages[[length(state$messages)]]
  result <- context_for_chat(ret, query, max_tokens = 1500, k = 3)
  list(context = result$context)
}

agent_node <- function(state, config) {
  context <- state$context %||% ""
  query <- state$messages[[length(state$messages)]]

  prompt <- paste0("Context:\n", context, "\n\nQuestion: ", query)
  chat <- chat_anthropic(
    system_prompt = "Answer questions about R packages using the provided context."
  )
  response <- chat$chat(prompt)

  # Persist what we learned
  mem$set(paste0("query.", Sys.time()), query)

  list(messages = list(response))
}

# --- Build and run ---
schema <- state_schema(messages = "append:list", context = "character")
g <- graph_builder(state_schema = schema)
g$add_node("retrieve", retrieve_node)
g$add_node("agent", agent_node)
g$add_edge("retrieve", "agent")
g$add_edge("agent", END)
g$set_entry_point("retrieve")

rag <- g$compile()
result <- rag$invoke(list(
  messages = list("How do I reshape data from wide to long format?")
))
```

The retrieval node finds chunks about tidyr's `pivot_longer`, the context
builder fits them within the token budget, and the agent answers using that
grounded context. The query is also persisted to the knowledge store for
future reference.
