---
title: "Retrieval Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retrieval Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Why Local Retrieval?

Retrieval-augmented generation (RAG) lets an LLM answer questions about your
own documents -- internal reports, package documentation, domain knowledge --
by fetching relevant passages and injecting them into the prompt. Most RAG
tooling requires sending your documents to an external embedding API, which
introduces latency, cost, and privacy concerns.

securecontext takes a **local-first** approach. Its TF-IDF embedder runs
entirely in your R process: no API keys, no network calls, no data leaving
your machine. This makes it suitable for air-gapped environments,
privacy-sensitive workloads, and rapid prototyping where you want results
without configuring external services.

This vignette walks through the complete retrieval pipeline step by step.
Every operation -- document creation, chunking, embedding, vector search,
knowledge storage, and context assembly -- runs locally.

## The Full RAG Pipeline

The following diagram shows how data flows from raw documents to an LLM-ready
context string:

```
  Ingest                      Search                     Assemble
  ------                      ------                     --------
  document()                  retrieve()                 context_builder()
      |                           |                           |
      v                           v                           v
  chunk_text()              embed query               cb_add() per chunk
      |                      with TF-IDF                      |
      v                           |                           v
  embed_tfidf()              cosine similarity          cb_build()
      |                      against store                    |
      v                           |                           v
  vector_store$add()         ranked results             token-limited
                                                        context string
```

The left column (ingest) happens once when you load your corpus. The middle
column (search) runs on every query. The right column (assemble) packs
results into a token budget before sending to an LLM.

```{r setup}
library(securecontext)
```

## Step 1: Create Documents

A `document()` wraps raw text with metadata and an auto-generated identifier.
Metadata travels with the document through the pipeline, making it easy to
trace which source a retrieved chunk came from. Documents are S7 objects, so
properties are accessed with `@`.

```{r documents}
doc_r <- document(
  "R is a programming language for statistical computing and graphics.
It is widely used among statisticians and data scientists. R provides
a wide variety of statistical and graphical techniques, including
linear and nonlinear modelling, classical statistical tests, time-series
analysis, classification, and clustering.",
  metadata = list(source = "intro", topic = "R")
)

doc_python <- document(
  "Python is a high-level, general-purpose programming language.
Its design philosophy emphasizes code readability. Python supports
multiple programming paradigms, including structured, object-oriented,
and functional programming. It is often used for machine learning,
web development, and data analysis.",
  metadata = list(source = "intro", topic = "Python")
)

doc_julia <- document(
  "Julia is a high-level, high-performance programming language for
technical computing. It was designed for numerical analysis and
computational science. Julia features optional typing, multiple dispatch,
and good performance approaching that of statically-typed languages
like C and Fortran.",
  metadata = list(source = "intro", topic = "Julia")
)

# S7 property access with @
doc_r@id
doc_r@metadata
```

## Step 2: Chunk Text

Chunking splits long text into smaller pieces suitable for embedding and
retrieval. Smaller, focused chunks improve search precision because the
embedder can match a query against a specific passage rather than an entire
document. The tradeoff is that very small chunks may lose surrounding context.

securecontext offers four strategies:

| Strategy      | How it splits                           |
|:------------- |:--------------------------------------- |
| `"sentence"`  | On sentence boundaries (`.` + space)    |
| `"paragraph"` | On double newlines                      |
| `"fixed"`     | Fixed character width with overlap       |
| `"recursive"` | Hierarchical separators (LangChain-style)|

### Choosing a Chunking Strategy

The right strategy depends on your content and how users will query it:

- **Sentence chunking** works best for narrative text (documentation, articles)
  where individual statements carry meaning. Queries like "What does dplyr do?"
  match well against single-sentence chunks.

- **Paragraph chunking** preserves more context per chunk, which helps when
  meaning spans multiple sentences. It works well for structured documents with
  clear paragraph breaks.

- **Fixed-size chunking** guarantees uniform chunk lengths, which is useful
  when downstream components expect consistent input sizes. The `overlap`
  parameter ensures that information near chunk boundaries is not lost.

- **Recursive chunking** is the most robust general-purpose strategy. It tries
  larger separators first (double newlines, then single newlines, then spaces,
  then characters), producing natural-looking chunks that respect document
  structure. This is a good default when you are unsure.

```{r chunking}
# Sentence-level chunking
sentences <- chunk_text(doc_r@text, strategy = "sentence")
cat("Sentence chunks:", length(sentences), "\n")
sentences

# Recursive chunking with a small max_size to demonstrate splitting
small_chunks <- chunk_text(doc_r@text, strategy = "recursive", max_size = 120)
cat("\nRecursive chunks (max 120 chars):", length(small_chunks), "\n")
small_chunks
```

Fixed-size chunking is useful when you need consistent chunk lengths, for
example when working with models that expect uniform input sizes. The `overlap`
parameter creates a sliding window, ensuring that information near chunk
boundaries appears in both adjacent chunks:

```{r fixed-chunking}
long_text <- paste(
  "The tidyverse is a collection of R packages designed for data science.",
  "It includes ggplot2 for visualization, dplyr for data manipulation,",
  "tidyr for data tidying, readr for reading data, purrr for functional",
  "programming, tibble for modern data frames, stringr for string",
  "manipulation, and forcats for factor handling.",
  "All tidyverse packages share an underlying design philosophy,",
  "grammar, and data structures."
)

fixed_chunks <- chunk_fixed(long_text, size = 100, overlap = 20)
cat("Fixed chunks (size=100, overlap=20):", length(fixed_chunks), "\n\n")
for (i in seq_along(fixed_chunks)) {
  cat(sprintf("Chunk %d (%d chars): %s\n\n", i, nchar(fixed_chunks[i]), fixed_chunks[i]))
}
```

## Step 3: Build a TF-IDF Embedder

Embeddings are numerical representations of text that capture semantic
similarity. Texts about similar topics produce vectors that are close together
in the embedding space, enabling search by meaning rather than exact keywords.

`embed_tfidf()` builds a vocabulary from a corpus and returns an embedder
that can project new texts into that TF-IDF space. TF-IDF (Term
Frequency-Inverse Document Frequency) weighs words by how important they
are to a specific document relative to the corpus. Common words like "the"
get low weights; distinctive words like "regression" get high weights.

Everything runs locally -- no API keys required.

```{r embedder}
# Gather all document texts as the training corpus
corpus <- c(doc_r@text, doc_python@text, doc_julia@text)

embedder <- embed_tfidf(corpus)

# The embedder is an S7 object; dims equals the vocabulary size
cat("Embedding dimensions:", embedder@dims, "\n")

# Embed a new query
query_matrix <- embed_texts(embedder, "statistical analysis in R")
cat("Query embedding shape:", nrow(query_matrix), "x", ncol(query_matrix), "\n")
```

## Step 4: Vector Store

The `vector_store` is an R6 class providing in-memory cosine-similarity search
with optional RDS persistence. It stores embedding vectors keyed by ID and
retrieves the closest matches to a query vector. Since it is R6, use `$` for
method access.

Cosine similarity measures the angle between two vectors, ignoring magnitude.
A score of 1.0 means identical direction (maximum similarity); 0.0 means
orthogonal (no similarity).

```{r vector-store}
vs <- vector_store$new(dims = embedder@dims)

# Embed and store each document manually
ids <- c("r", "python", "julia")
embeddings <- embed_texts(embedder, corpus)
vs$add(ids, embeddings)

cat("Store size:", vs$size(), "vectors\n")

# Search for the closest match to a query
query_emb <- embed_texts(embedder, "data science and statistics")
results <- vs$search(query_emb, k = 3)
print(results)
```

Persistence is straightforward with `$save()` and `$load()`. This lets you
build an embedding index once and reuse it across R sessions without
re-embedding your corpus:

```{r persistence}
tmp <- tempfile(fileext = ".rds")
vs$save(tmp)

vs2 <- vector_store$new(dims = embedder@dims)
vs2$load(tmp)
cat("Loaded store size:", vs2$size(), "\n")

# Clean up
unlink(tmp)
```

## Step 5: Retriever -- the High-Level Interface

The previous steps (chunk, embed, store, search) are the building blocks.
The `retriever()` combines a vector store and an embedder into a single object
that handles the full ingest-and-search workflow. Use `add_documents()` to
chunk, embed, and store documents in one call, then `retrieve()` to search.

This is the recommended interface for most applications -- it reduces
boilerplate and ensures that chunking and embedding are consistent between
ingest and query time.

```{r retriever}
# Fresh store for the retriever
vs_ret <- vector_store$new(dims = embedder@dims)
ret <- retriever(vs_ret, embedder)

# add_documents handles chunking + embedding internally
docs <- list(doc_r, doc_python, doc_julia)
add_documents(ret, docs, chunk_strategy = "sentence")

cat("Chunks in store:", vs_ret$size(), "\n\n")

# Retrieve the top 3 chunks for a query
hits <- retrieve(ret, "machine learning", k = 3)
print(hits)
```

The returned data frame contains chunk IDs and cosine similarity scores. Higher
scores indicate greater relevance. You can use these scores directly as
priorities in the context builder (see `vignette("context-building")`).

## Step 6: Knowledge Store

The `knowledge_store` is an R6 class providing persistent key-value storage
backed by JSONL. While the vector store is optimized for similarity search
over embeddings, the knowledge store is designed for structured data you look
up by key: agent memory, user preferences, session history, learned facts.

The JSONL format (one JSON object per line) makes the store append-friendly
and human-readable. Data persists across R sessions automatically.

```{r knowledge-store}
ks <- knowledge_store$new(path = tempfile(fileext = ".jsonl"))

# Store some facts
ks$set("lang.r", list(type = "statistical", year = 1993))
ks$set("lang.python", list(type = "general-purpose", year = 1991))
ks$set("lang.julia", list(type = "numerical", year = 2012))
ks$set("user.preference", "R")

cat("Total entries:", ks$size(), "\n")

# Retrieve a value
ks$get("lang.r")

# Search keys by regex
ks$search("^lang")

# List all keys
ks$list()

# Clean up
unlink(ks$.__enclos_env__$private$.path)
```

## Step 7: Context Builder

The `context_builder()` assembles a token-limited context string from multiple
sources, prioritizing the most important content. This is the final step before
sending context to an LLM -- it ensures you stay within the model's context
window while including the most relevant information.

For a deep dive into priority strategies, overflow behavior, and multi-turn
patterns, see `vignette("context-building")`.

```{r context-builder}
cb <- context_builder(max_tokens = 100)

# Add content with different priorities (higher = included first)
cb <- cb_add(cb, "You are a helpful assistant.", priority = 10, label = "system")
cb <- cb_add(cb,
  "R is great for statistics and data visualization.",
  priority = 5, label = "retrieved_chunk_1"
)
cb <- cb_add(cb,
  "Python is popular for machine learning and web development.",
  priority = 4, label = "retrieved_chunk_2"
)
cb <- cb_add(cb,
  "Julia offers high performance for numerical computing workloads.",
  priority = 3, label = "retrieved_chunk_3"
)

result <- cb_build(cb)

cat("Assembled context:\n")
cat(result$context, "\n\n")
cat("Included:", paste(result$included, collapse = ", "), "\n")
cat("Excluded:", paste(result$excluded, collapse = ", "), "\n")
cat("Total tokens:", result$total_tokens, "\n")
```

The builder processes items in priority order and stops adding when the token
budget is exhausted. Items that do not fit are reported in `$excluded`, making
it easy to log what was dropped.

Use `cb_reset()` to clear all items and reuse the same builder with a new turn
of conversation:

```{r context-reset}
cb2 <- cb_reset(cb)
cb2 <- cb_add(cb2, "New system prompt.", priority = 10, label = "system_v2")
result2 <- cb_build(cb2)
cat("After reset -- included:", paste(result2$included, collapse = ", "), "\n")
```

## Full Pipeline: Retrieve and Build Context

`context_for_chat()` combines retrieval and context building in a single call.
Given a retriever and a query, it retrieves the top-k chunks and packs them
into a token-limited context string. This is the typical integration point for
an agent: retrieve relevant information, assemble a context window, and pass
it to an LLM provider.

```{r full-pipeline}
context_result <- context_for_chat(ret, "statistical computing", max_tokens = 2000, k = 5)

cat("Context for LLM:\n")
cat(context_result$context, "\n\n")
cat("Included chunks:", length(context_result$included), "\n")
cat("Total tokens:", context_result$total_tokens, "\n")
```

## Summary

The securecontext retrieval pipeline follows these steps:

1. **`document()`** -- wrap text with metadata
2. **`chunk_text()`** -- split into retrieval units
3. **`embed_tfidf()`** -- build a local embedder from a corpus
4. **`vector_store$new()`** -- store and search embeddings
5. **`retriever()`** + **`add_documents()`** -- high-level ingest
6. **`retrieve()`** -- semantic search
7. **`knowledge_store$new()`** -- persistent key-value memory
8. **`context_builder()`** + **`cb_add()`** + **`cb_build()`** -- token-aware assembly
9. **`context_for_chat()`** -- one-call retrieve-and-build

For more on context building strategies, see `vignette("context-building")`.
To wire retrieval into orchestr agent graphs, see
`vignette("orchestr-integration")`.
