---
title: "Getting Started with securecontext"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with securecontext}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

securecontext provides memory, knowledge persistence, RAG retrieval, and
token-aware context management for R LLM agents.

## Documents and Chunking

Create documents and split them into chunks:

```{r}
library(securecontext)

doc <- document(
  "R is a language for statistical computing. It has many packages for data
analysis. The tidyverse is a popular collection of packages.",
  metadata = list(source = "intro")
)

# Different chunking strategies
chunks_sent <- chunk_text(doc$text, strategy = "sentence")
chunks_para <- chunk_text(doc$text, strategy = "paragraph")
chunks_rec  <- chunk_text(doc$text, strategy = "recursive", max_size = 50)
```

## Embeddings and Vector Search

Build TF-IDF embeddings locally (no API needed):

```{r}
corpus <- c(
  "R is great for statistics and data analysis.",
  "Python excels at machine learning and deep learning.",
  "Julia is fast for numerical computing."
)

embedder <- embed_tfidf(corpus)
vs <- vector_store$new(dims = embedder$dims)

# Add documents
docs <- lapply(corpus, document)
ret <- retriever(vs, embedder)
add_documents(ret, docs)

# Search
results <- retrieve(ret, "statistical computing", k = 2)
print(results)
```

## Knowledge Store

Persistent key-value storage backed by JSONL:

```{r}
ks <- knowledge_store$new(path = tempfile(fileext = ".jsonl"))

ks$set("user.name", "Alice")
ks$set("user.preferences", list(theme = "dark", lang = "R"))

ks$get("user.name")
ks$search("^user")
```

## Context Builder

Assemble token-limited context with priorities:

```{r}
cb <- context_builder(max_tokens = 500)
cb <- cb_add(cb, "System instructions go here.", priority = 10, label = "system")
cb <- cb_add(cb, "Retrieved context from documents.", priority = 5, label = "context")
cb <- cb_add(cb, "Chat history summary.", priority = 3, label = "history")

result <- cb_build(cb)
cat(result$context)
result$included
result$excluded
```

## Integration

Use `context_for_chat()` to combine retrieval and context building:

```{r}
result <- context_for_chat(ret, "statistics", max_tokens = 2000)
cat(result$context)
```

Wrap a knowledge store for orchestr:

```{r}
mem <- as_orchestr_memory(ks)
mem$set("key", "value")
mem$get("key")
```
